Wine Quality Reds by Heidi Raasch
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#loading the libraries
library(ggplot2)
library(dplyr)
library(GGally)
library(RColorBrewer)
library(reshape)
library(gridExtra)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
#Change working directory
setwd("C:/Users/Heidi/Documents/Explore and Summarize Data")
# Load the Data
wine <- read.csv('wineQualityReds.csv')

```

This report explores a dataset containing information on the relationship
between red wine attributes and quality. 
 

# Univariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
# First we should take a glimpse into our data to see what we're working with.
glimpse(wine)

```

X appears to be a index for each row of the data. Therefore, we will exclude 
that variable going foward. 

#Examine histograms of all variables

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}
#I used the reshape library to melt the wide data and then visualize the data 
#distribution for each variable using boxplots.  

meltData <- melt(wine[2:13])

ggplot(meltData, aes(x = value)) + 
    facet_wrap(~variable,scales = "free_x") + 
    geom_histogram()


```

We can see that most of the variables have a normal distribution. There are 
outliers in the residual.sugar, sulphates and chlorides variables that make it 
difficult to properly view the distribution. To make the distributions of these 
3 variables easier to see we will exclude the 95 percentile.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}
#creating a plot of the residual sugar, excluding the 95th percentile
p1 <- ggplot(wine, aes(x=residual.sugar)) + 
  geom_histogram() + 
  xlim(0, quantile(wine$residual.sugar, c(0.95)))    
#creating a plot of the chlorides, excluding the 95th percentile
p2 <- ggplot(wine, aes(x=chlorides)) + 
  geom_histogram() + 
  xlim(0, quantile(wine$chlorides, c(0.95)))
#creating a plot of the sulphates, excluding the 95th percentile
p3 <- ggplot(wine, aes(x=sulphates)) + 
  geom_histogram() + 
  xlim(0, quantile(wine$sulphates, c(0.95)))
#arranging the grids to appear on one row with 3 columns
grid.arrange(p1, p2, p3, ncol=3)

```

We can also see that there is a problem with the quality plot. We do not want
there to be spaces between the bars on the hisograms as that can indicate a
possiblitity of observations between the bars. Therefore we will use a bar plot 
instead of the histogram. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2.1}
#I used the reshape library to melt the wide data and then visualize the data 
#distribution for each variable using boxplots.  

meltData <- melt(wine[2:13])

ggplot(data = wine, aes(x = quality)) +
  geom_histogram(binwidth = 1)


```

After removing the outliers we can see the distributions of residual.sugar, 
chlorides, and sulphates are also normal. 

I am most interested in the quality of the wines and what variables have an 
impact on the quality.

#Quality
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}
ggplot(data = wine, aes(x = quality)) +
  geom_bar(binwidth = 1)

```

The majority of the wines are of average quality (4.5-6.5) and there are few 
wines with low or high quality. The lowest quality is 3 and the highest quality
is 8. 

#PH
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}
#Create a histogram of the pH of the wines

ggplot(data = wine, aes(x = pH)) +
  geom_histogram()

```

The distribution of the pH of the wines is similar to the distribution of the 
quality of the wines. It is a normal distribution with a majority of the wines 
having a pH between 3.0 and 3.7 with a few wines having a pH higher than 3.7 or
lower than 3.0.

#Alcohol

Alcohol is another important factor of red wines. I will explore that next.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots6}
#Create a histogram of the alcohol percentage of the wines

ggplot(data = wine, aes(x = alcohol)) +
  geom_histogram()

```

Unlike pH and qality we see that alcohol is right skewed. We see a 
majority of the wines have an alcohol content that falls on the left side of
the plot. 


# Univariate Analysis

### What is the structure of your dataset?

The dataset has 12 variables representing 1599 observations. Each observation 
corresponds to a red wine sample. 11 of the variables correspond to the result
of a physicochemical test and the last variable (quality) corresponds to the 
result of a sensory panel rating.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in this dataset is quality. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Each of the physicochemical variables will help support the investigation as 
they each can impact the taste and quality of the wine. Also, I expect that 
some of the variables will have a high correlation to each other. 

### Did you create any new variables from existing variables in the dataset?

I created the meltData variable which made use of the reshape library to melt 
the wide data.  


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Originally the residual.sugar, sulphates and chlorides variables had skewed
distributions that contained outliers. I excluded the 95th percentile to remove
the outliers which resulted in normal distributions. 


# Bivariate Plots Section

```{r results, message=FALSE, warning=FALSE, include=FALSE}
#Generating the correlation between each of the variables and quality

with(wine, cor.test(quality, fixed.acidity))
with(wine, cor.test(quality, volatile.acidity))
with(wine, cor.test(quality, citric.acid))
with(wine, cor.test(quality, residual.sugar))
with(wine, cor.test(quality, chlorides))
with(wine, cor.test(quality, free.sulfur.dioxide))
with(wine, cor.test(quality, density))
with(wine, cor.test(quality,  pH))
with(wine, cor.test(quality, sulphates))
with(wine, cor.test(quality, alcohol))

```

The correlations between each of the variables and quality are:
Fixed Acidity: 0.1240516
Volatile Acidity: -0.3905578
Citric Acid: 0.2263725
Residual Sugar: 0.01373164 
Chlorides: -0.1289066 
Free Sulfur Dioxide:-0.05065606 
Density: -0.1749192  
pH: -0.05773139 
Sulphates: 0.2513971 
Alcohol: 0.4761663

The top 4 correlation coefficents to the quality of the wine are:
  Alcohol: 0.48
  Sulphates: 0.25
  Citric Acid: 0.23
  Fixed Acidity: 0.12

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots2}
#Creating plots of the top 4 correlation coefficents
#Plotting the alcohol and quality coefficent
p1 <- ggplot(wine, aes(x = alcohol, y = quality)) + 
    geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'red')

#Plotting the sulphates and quality coefficent
p2 <- ggplot(wine, aes(x = sulphates, y = quality)) + 
    geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'red')

#Plotting the citric acid and quality coefficent
p3 <- ggplot(wine, aes(x = citric.acid, y = quality)) + 
    geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'red')

#Plotting the fixed acidity and quality coefficent
p4 <- ggplot(wine, aes(x = fixed.acidity, y = quality)) + 
    geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'red')
#Arranging the plots in a 2 by 2 grid
grid.arrange(p1, p2, p3, p4, ncol=2, nrow=2)

```

We can see that the quality of the wine increases with as the alcohol, 
sulphates, citric acid, and fixed acidity increase. 

Lets take a look at some of the correlations between alcohol and some of the
other variables. 

```{r results2, message=FALSE, warning=FALSE, include=FALSE}
#Generating the coefficents between alcohol and each of the variables
with(wine, cor.test(alcohol, fixed.acidity))
with(wine, cor.test(alcohol, volatile.acidity))
with(wine, cor.test(alcohol, citric.acid))
with(wine, cor.test(alcohol, residual.sugar))
with(wine, cor.test(alcohol, chlorides))
with(wine, cor.test(alcohol, free.sulfur.dioxide))
with(wine, cor.test(alcohol, density))
with(wine, cor.test(alcohol,  pH))
with(wine, cor.test(alcohol, sulphates))

```

The correlations between each of the variables and alcohol are:
Fixed Acidity: -0.06166827 
Volatile Acidity: -0.202288 
Citric Acid: 0.1099032 
Residual Sugar: 0.04207544 
Chlorides: -0.2211405
Free Sulfur Dioxide: -0.06940835
Density: -0.4961798   
pH: 0.2056325 
Sulphates: 0.09359475 

The variables with the highest (positive or negative) correlations are:
  Density: -0.50
  Chlorides: -0.22
  pH: 0.21
  Volatile Acidity: -0.20

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots4}
#Plotting the alcohol and quality coefficent 
p1 <- ggplot(wine, aes(x = density, y = alcohol)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'blue')
#Plotting the alcohol and quality coefficent
p2 <- ggplot(wine, aes(x = chlorides, y = alcohol)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'blue')
#Plotting the alcohol and quality coefficent
p3 <- ggplot(wine, aes(x = pH , y = alcohol)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'blue')
#Plotting the alcohol and quality coefficent
p4 <- ggplot(wine, aes(x = volatile.acidity, y = alcohol)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'blue')
#Arrange the plots into a 2 by 2 grid
grid.arrange(p1, p2, p3, p4, ncol=2, nrow=2)

```

We can see that the density, chlorides, and volatile acidity of the wine 
decrease with the amount of alcohol in the wine. Whereas the pH tends to 
increase with the amount of alcohol. 

Lets explore the correlation between pH, citric acid, volatile acidity, and
fixed acidity. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots5}
#Plotting the citric acid and pH coefficent 
p1 <- ggplot(wine, aes(x = citric.acid, y = pH)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'green')

#Plotting the volatile acidity and pH coefficent 
p2 <- ggplot(wine, aes(x = volatile.acidity, y = pH)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'green')

#Plotting the fixed acidity and pH coefficent 
p3 <- ggplot(wine, aes(x = fixed.acidity , y = pH)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  geom_smooth(method = 'lm', color = 'green')

#Arranging the plots in a 2 by 2 grid
grid.arrange(p1, p2, p3, nrow= 2, ncol=2)

```

Here we can see that the pH decreases as the citric acid and fixed acidity 
increase whereas the pH increases as the volatile acidity increases. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The quality of the wine increases with as the alcohol, sulphates, citric acid, 
and fixed acidity increase.

The density, chlorides, and volatile acidity of the wine 
decrease with the amount of alcohol in the wine. Whereas the pH tends to 
increase with the amount of alcohol. 

The pH decreases as the citric acid and fixed acidity increase whereas the 
pH increases as the volatile acidity increases.

### Did you observe any interesting relationships between the other features \ 
(not the main feature(s) of interest)?

I found it interesting that the density decreases with the amount of alcohol 
in the wine.

### What was the strongest relationship you found?

The strongest relationship was between the quality and alcohol in the wine. 

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
#Create a plot displaying the relationships between volatile acidity, pH, 
#alcohol, and quality.
ggplot(wine, aes(x=volatile.acidity, y=pH, size = alcohol, 
                 color=factor(quality))) + 
  geom_point() + 
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality")

```

We see that a higher concentration of the blue dots are located on the left 
side of the plot. This indicates the higher quality wines have lower volatile 
acidity. This is also where we see the larger dots representing the wines with 
a higher alcohol content. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}
#Plot the relationship between pH, alcohol, and quality using a contour plot
ggplot(wine, aes(x = pH, y = alcohol, 
                 colour=factor(wine$quality))) + 
  geom_density2d(bins=2) + 
   scale_fill_brewer(type = "div", palette = "RdYlBu", name="Quality") + 
  geom_point(color="black", alpha=0.1) 



```

Here we see the higher quality wines represented by the blue, teal and pink 
lines. We see they lay close to the center of the plot. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}
#Create a plot that shows the quality in density plots and color.
 
empty <- ggplot()+geom_point(aes(1,1), colour="white") +
     theme( 
       #remove background color
       plot.background = element_blank(),
       #remove grid lines
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank(), 
       #remove border
       panel.border = element_blank(), 
       #Remove the panel background color
       panel.background = element_blank(),
       #Remove title, text, and tick marks
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
       axis.text.y = element_blank(),
       axis.ticks = element_blank()
     )
#Create the main plot
main <- ggplot(wine, aes(x = pH, y = alcohol, 
                 color=factor(wine$quality))) + 
  geom_point() +
  theme(legend.position=c(1,1),legend.justification=c(1,1)) +
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality") 

#Create the density plot for ph and quality
top <- ggplot(wine, aes(x = pH, 
                 fill=factor(wine$quality))) + 
  geom_density(alpha=0.5) +
  theme(legend.position = "none") +
   scale_fill_brewer(type = "div", palette = "RdYlBu", name="Quality")
#Create the density plot for alcohol and quality  
right <- ggplot(wine, aes(x = alcohol, 
                 fill=factor(wine$quality))) + 
  coord_flip() + 
  geom_density(alpha=0.5) + 
  theme(legend.position = "none") +
   scale_fill_brewer(type = "div", palette = "RdYlBu", name="Quality")
#Arrange the plots

grid.arrange(top, empty, main, right, ncol=2, nrow=2, widths=c(4, 1),
             heights=c(1, 4))
```

Again we see that the higher quality wines are concentrated more in the center
of the plot.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The main relationshps explored were between the pH and alcohol content. 

Wines that have a mid-range pH and alcohol conent tend to have higher quality.


### Were there any interesting or surprising interactions between features?

The relationship between pH and alcohol is interesting.


------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
#Plot the relationship between quality and alcohol 
ggplot(wine, aes(x = alcohol, y = quality)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  ylab("Quality Rating")+
  xlab("Alcohol (% by Volume)")+
  geom_smooth(method = 'lm', color = 'red')+
  ggtitle("Higher alcohol, higher quality")+
  theme(plot.title = element_text(hjust = 0.5, size = 20), 
        text = element_text(size = 12))

```

### Description One
This plot demonstrates how the quality of a wine increases as the alcohol 
content increase. We see a high concentration of the points in the middle
of the quality axis.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
#Plot the relationship between pH and alcohol
ggplot(wine, aes(x = pH , y = alcohol)) + 
  geom_jitter(alpha=0.5, width = 0, height = 0.25) +
  ylab("Alcohol (% by Volume)")+
  xlab("pH")+
  geom_smooth(method = 'lm', color = 'blue')+
  ggtitle("Alcohol and pH")+
  theme(plot.title = element_text(hjust = 0.5, size = 20), 
        text = element_text(size = 12))

```

### Description Two
This plot shows the relationship between pH and Alcohol. We see that the pH 
slightly increases as the alcohol content gets higher. We can also see how the
points are more spread out as the alcohol content gets higher and they get more
condensed the lower the alcohol content is. 

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
#Create a plot that shows the quality in density plots and color.
 
empty <- ggplot()+geom_point(aes(1,1), colour="white") +
     theme( 
       #remove background color
       plot.background = element_blank(),
       #remove grid lines
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank(), 
       #remove border
       panel.border = element_blank(), 
       #Remove the panel background color
       panel.background = element_blank(),
       #Remove title, text, and tick marks
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
       axis.text.y = element_blank(),
       axis.ticks = element_blank()
     )

#Create the plot of the relationship between pH, alcohol, and quality
ggplot(wine, aes(x = pH, y = alcohol, color=factor(wine$quality))) + 
  geom_point() +
  ylab("Alcohol (% by Volume)")+
  xlab("pH")+
  ggtitle("Alcohol, pH, and Quality")+
  theme(plot.title = element_text(hjust = 0.5, size = 20), 
        text = element_text(size = 12),
        legend.position=c(1,1),legend.justification=c(1,1))+
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality") 



```

### Description Three
This plot shows the relationship between the alcohol, pH, and quality of the 
wines. We see the higher quality wines represented by the blue dots. 
The highest density of those colors tends to be in the center of each of the 
variables. 

We also see a majority of the dots on the plot are orange and red. This 
demonstrates what we saw in the Univariate exploration.

# Reflection

This project was interesting as I was able to put my new found knowledge of
the plotting functionality of R to the test on a real dataset. There are 
several ways I could have gone with this data and I found it difficult trying 
to narrow down my exploration. This dataset was already clean which made it 
easy to work with. 

Future work with this dataset could include combining the white wine dataset
with this dataset to find if there are differences when it comes to red vs
white wines. 
